\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-05 misc/JSS article/final/article.tex
Further fixes to join()/pivot() section.

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations provide outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
  drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+                   labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{If multiple columns are selected, they are combined using \code{"\_"} for names and \code{" - "} for labels.} Note below how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation---with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{drop}, \code{sort}, and \code{transpose} can be used to control the casting process/output.
%
\begin{Schunk}
\begin{Sinput}
R> head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
+                   labels = "Label", how = "w"))
\end{Sinput}
\begin{Soutput}
  Country Year AGR_VA AGR_EMP MIN_VA MIN_EMP MAN_VA MAN_EMP
1     BWA 1964  16.30   152.1  3.494  1.9400 0.7366   2.420
2     BWA 1965  15.73   153.3  2.496  1.3263 1.0182   2.330
3     BWA 1966  17.68   153.9  1.970  1.0022 0.8038   1.282
4     BWA 1967  19.15   155.1  2.299  1.1192 0.9378   1.042
5     BWA 1968  21.10   156.2  1.839  0.7855 0.7503   1.069
6     BWA 1969  21.86   157.4  5.245  2.0314 2.1396   2.124
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                       Label
1  Country                     Country
2     Year                        Year
3   AGR_VA   Agriculture - Value Added
4  AGR_EMP    Agriculture - Employment
5   MIN_VA        Mining - Value Added
6  MIN_EMP         Mining - Employment
7   MAN_VA Manufacturing - Value Added
8  MAN_EMP  Manufacturing - Employment
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names needs to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} element will omit the respective operation.
%
\begin{Schunk}
\begin{Sinput}
R> head(dr <- pivot(data, c("Country", "Year"),
+               names = list(from = "Variable", to = "Sectorcode"),
+               labels = list(from = "Label", to = "Sector"), how = "r"))
\end{Sinput}
\begin{Soutput}
  Country Year Sectorcode      Sector    VA   EMP
1     BWA 1964        AGR Agriculture 16.30 152.1
2     BWA 1965        AGR Agriculture 15.73 153.3
3     BWA 1966        AGR Agriculture 17.68 153.9
4     BWA 1967        AGR Agriculture 19.15 155.1
5     BWA 1968        AGR Agriculture 21.10 156.2
6     BWA 1969        AGR Agriculture 21.86 157.4
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)[3:6]
\end{Sinput}
\begin{Soutput}
   Sectorcode        Sector            VA           EMP
           NA            NA "Value Added"  "Employment"
\end{Soutput}
\end{Schunk}
%
This (\code{dr}) is the tidy format \citep{rtidydata} where each variable is a separate column. It is analytically more useful, e.g., to compute labor productivity as \code{settransform(dr, LP = VA / EMP)} or to estimate a panel-regression with sector fixed-effects. The recast pivot is thus a natural operation to change data representation. As with the other pivots, it preserves all information and can be reversed by simply swapping the contents of the \code{from} and \code{to} keywords. \newline

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
+       names = list(from = "Variable", to = "Sectorcode"),
+       labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

>>>>

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations provide outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+                   labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{If multiple columns are selected, they are combined using \code{"\_"} for names and \code{" - "} for labels.} Note below how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation---with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{drop}, \code{sort}, and \code{transpose} can be used to control the casting process/output.
%
\begin{Schunk}
\begin{Sinput}
R> head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
+                   labels = "Label", how = "wider"))
\end{Sinput}
\begin{Soutput}
  Country Year AGR_VA AGR_EMP MIN_VA MIN_EMP MAN_VA MAN_EMP
1     BWA 1964  16.30   152.1  3.494  1.9400 0.7366   2.420
2     BWA 1965  15.73   153.3  2.496  1.3263 1.0182   2.330
3     BWA 1966  17.68   153.9  1.970  1.0022 0.8038   1.282
4     BWA 1967  19.15   155.1  2.299  1.1192 0.9378   1.042
5     BWA 1968  21.10   156.2  1.839  0.7855 0.7503   1.069
6     BWA 1969  21.86   157.4  5.245  2.0314 2.1396   2.124
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                       Label
1  Country                     Country
2     Year                        Year
3   AGR_VA   Agriculture - Value Added
4  AGR_EMP    Agriculture - Employment
5   MIN_VA        Mining - Value Added
6  MIN_EMP         Mining - Employment
7   MAN_VA Manufacturing - Value Added
8  MAN_EMP  Manufacturing - Employment
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names needs to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} element will omit the respective operation.
%
\begin{Schunk}
\begin{Sinput}
R> head(dr <- pivot(data, c("Country", "Year"),
+    names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), how = "recast"))
\end{Sinput}
\begin{Soutput}
  Country Year Sectorcode      Sector    VA   EMP
1     BWA 1964        AGR Agriculture 16.30 152.1
2     BWA 1965        AGR Agriculture 15.73 153.3
3     BWA 1966        AGR Agriculture 17.68 153.9
4     BWA 1967        AGR Agriculture 19.15 155.1
5     BWA 1968        AGR Agriculture 21.10 156.2
6     BWA 1969        AGR Agriculture 21.86 157.4
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)[3:6]
\end{Sinput}
\begin{Soutput}
   Sectorcode        Sector            VA           EMP
           NA            NA "Value Added"  "Employment"
\end{Soutput}
\end{Schunk}
%
This (\code{dr}) is the tidy format \citep{rtidydata} where each variable is a separate column. It is analytically more useful, e.g., to compute labor productivity as \code{settransform(dr, LP = VA / EMP)} or to estimate a panel-regression with sector fixed-effects. The recast pivot is thus a natural operation to change data representation. As with the other pivots, it preserves all information and can be reversed by simply swapping the contents of the \code{from} and \code{to} keywords. \newline

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline
